Depression and suicidality as costly signals of need

### 2021-04-01 Paper.rmd
Started to work on the intro.

# Introduction
Since Durkheim’s pioneering effort to use quantitative social science to explain suicide (cite), anthropologists, psychologists, and sociologists have made varied attempts at better understanding depression and suicide. Despite this effort and substantial investment by funding agencies, rates of depression and suicide attempts have remained largely unchanged over the last 100 years (confirm and cite), and there is still substantial uncertainty about why both are as common as they are (cite). Likely due to  the suffering they cause, the vast majority of these approaches have been built on the assumption that they are the result of internal dysfunction (cite examples). However, a subset of anthropologists and evolutionary theorists have challenged this idea, suggesting that despite the subjective suffering they cause, they may function as a means to limit negative fitness outcomes in times of substantial adversity (Firth + others).

Within anthropology, Raymond Firth was likely the first to do this. Based on observations that suicide attempts often followed social conflict or loss and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, and immediate reintegration into the community following negative events (Firth 1967[not original date]). Compared to suicide, however, anthropological explanations for depression are a much more recent occurrence (cite), despite frequent reports of depression or depression-like behavior in a wide range of societies (cite) and the potential for cross-cultural research to help understand depressive states (cite).

## Connection to Adversity
Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005 + others; citations for suicide). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, it's onset tends to quickly follow, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

>>>>

# Introduction

Durkheim’s pioneering study of suicide [@durkheim1897] arguably launched sociology as a quantitative discipline. Sociological investigations of suicide flourished through much of the 20th century, but began to be supplanted by medical and public health approaches as the biomedical model of mental health took hold in the late 1970's and early 1980's [@wray2011; @syme2020]. The biomedical model tends to frame suicide as an individiual-level problem, e.g., as one involving psychiatric disturbance, drug addiction, or severe stress [@wray2011]. This theoretical approach has made little progress: Despite substantial investment by funding agencies, rates of suicide have remained largely unchanged over the last 100 years (confirm and cite).

Early on, anthropologists also tackled the problem of suicide as manifested in the small, traditional socities in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal (e.g., Malinowski, 1985; Niehaus, 2012). Some ethnographers, such as Hezel (1987), conceptualized suicide as a form of anger, whereas others, such as Giddens (1964), conceptualized suicide as a form of social pressure. Counts (1980) emphasized the powerlessness of suicide victims. Common to almost all theoretical and empirical investigations of suicide is a focus on *completed* suicides, i.e., suicide deaths. @SymeEtAl2016, in contrast, argued that because the vast majority of suicidal behavior does not result in death, the proper focus should be on suicide ideation and suicide attempts, not deaths.

Within anthropology, Raymond Firth was likely the first to put the focus on suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed social conflict or loss and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, and immediate reintegration into the community following negative events (Firth 1967[not original date]). Compared to suicide, however, anthropological explanations for depression are a much more recent occurrence (cite), despite frequent reports of depression or depression-like behavior in a wide range of societies (cite) and the potential for cross-cultural research to help understand depressive states (cite).

and there is still substantial uncertainty about why both are as common as they are (cite). Likely due to  the suffering they cause, the vast majority of these approaches have been built on the assumption that they are the result of internal dysfunction (cite examples). However, a subset of anthropologists and evolutionary theorists have challenged this idea, suggesting that despite the subjective suffering they cause, they may function as a means to limit negative fitness outcomes in times of substantial adversity (Firth + others).

## Connection to Adversity

Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005 + others; citations for suicide). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, it's onset tends to quickly follow, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).
